---
title: "Evaluation of Bridge Saftey in Omaha,NE"
author: "Tyler Davis"
bibliography: references.bib
format:
  revealjs:
    incremental: true
output-file: index.html
---

# Introducing the Problem

## Setting the Stage

-   Have you ever been driving over a bridge and wondering how stable it is?

-   The United States received a C rating on their infrastructure for 2025, @NebraskaInfrastructureReport2025

    -   The best the have received in years

## Major Collapses

-   In recent years the two major collapse in the United States have been:
    -   the Francis Scott Key bridge in Baltimore MD which collapsed last year
    -   I - 35 W Mississippi River Bridge located in Minneapolis, MN collapsed in 2007, @deletterBaltimoresKeyBridge2024

## Why it Matters

-   The National Transportation Safety Board flagged 68 bridges that are at risk of collapse from vessel strikes, @NTSBRecommends682025

-   The bridge condition in Nebraska is also in question as 8.8% of bridges are deficient

-   149 dams are considered high hazard potential.

## Problem Statement

-   What is probability of a bridge collapse in Omaha, NE?

## Starting Analysis

-   NDAT National Bridge Inventory data set, @NationalBridgeInventory2024
-   It contains 130 variables describing:
    -   bridge location, structural condition, and average traffic
-   Filtered to contain only the Omaha area, which includes towns like Elkhorn, NE and Council Bluffs, IA.

# Methods

## Graph of Omaha Bridges

::::: columns
::: {.column width="70%"}
```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(psych)
library(knitr)
options(digits = 4)
omaha_bridges <- read.csv("omaha_bridges.csv")


```

```{r}
#| label: fig-omaha-map1
#| fig-cap: "Map of Omaha Bridges"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
library(leaflet)
leaflet(omaha_bridges) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )
```
:::

::: {.column width="30%"}
-   This map displays the overall distribution of bridges in Omaha, NE
:::
:::::

## Understanding the Bridge Condition Variable

-   The Bridge Condition variable
    -   has rating of "P" for poor, "G" for good, and "F" for fair
-   Bridge condition rating is determined by:
    -   The lowest rating of condition ratings for:
        -   Deck, Superstructure, Substructure, and Culvert
-   If the lowest rating is greater than or equal to 7 = "G", if it is 5 or 6 the rating is "F", and if it is less than or equal to 4, the rating is "P"

## Bridge Condition Analysis

::::: columns
::: {.column width="60%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
fil_dat <- omaha_bridges %>%
  select(c(1,2,5,14,21,22,25, 28:29,31:32, 68:77,88,103,107,112,116, 122:125))
library(magrittr)
desc_dat <- data.frame(variable = names(fil_dat),
                       classes = sapply(fil_dat, typeof),
                       first_values = sapply(fil_dat, function(x) paste0(head(x),  collapse = ", ")),
                       row.names = NULL)
kable(desc_dat)

library(ggplot2)

```

```{r}
#| label: fig-omaha-brid
#| fig-cap: "Bridge Condtion Ratings Across Omaha, NE"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
ggplot(fil_dat,aes(x = BRIDGE_CONDITION)) + geom_bar() + labs(title ="Bridge Condition") + xlab("Bridge Rating")

fil_dat <- fil_dat %>%
  mutate(across(everything(), ~ ifelse(. == "N", NA, .)))

```
:::

::: {.column width="40%"}
-   The graph shows the overall breakdown of how the bridge condition variable.
-   This shows that a majority of the bridges in Omaha are in good or fair condition.
-   Next the bridges with a poor rating will be analyzed.
:::
:::::

## Bridges in Poor Condition with and without High Traffic

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-omaha-map
#| fig-cap: "Map of Omaha Bridges in 'Poor' Condtion"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
bridge_cond <- fil_dat %>%
  filter(
    BRIDGE_CONDITION == "P"
  ) %>%
  select(c(1,4,5,6,8,9:21,29,27,23))


leaflet(bridge_cond) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: fig-omaha-map2
#| fig-cap: "Map of Omaha Bridges in 'Poor' Condtion and with High Traffic"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
high_traffic <- bridge_cond %>%
  filter (
    ADT_029 > 5000 & TRAFFIC_LANES_ON_028A > 2
  )
leaflet(high_traffic) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ LONG_017,
    lat = ~ LAT_016,
    radius = 5,
    color = "red",
    popup = ~paste("Bridge ID:", FACILITY_CARRIED_007)
  )

```
:::
:::::

# Analysis

## Descriptive Statistics: Bridges with Poor Condition

::::: columns
::: {.column width="60%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
library(psych)
bridge_cond1 <- fil_dat %>%
 filter(
   BRIDGE_CONDITION == "P"
 ) %>%
   select(c(8,9,10,12:16,21,26,27))

high_traffic <- bridge_cond1 %>%
  filter (
    ADT_029 > 5000 & TRAFFIC_LANES_ON_028A > 2
  )

desc_brid_cond <- describe(bridge_cond1)
desc_h_t <- describe(high_traffic)

```

```{r}

#| label: tbl-bridge-stats
#| tbl-cap: "Descriptive statistics of bridges in poor condition"
#| tbl-width: 6
#| tbl-height: 4
#| echo: false
#| message: false


library(kableExtra)

desc_brid_cond %>%
  select(mean, sd, n, median) %>%
  mutate(across(c(mean, sd, median), ~ format(round(.x, 3), scientific = FALSE))) %>%
  kable() %>%
  kable_styling(font_size = 17) 
```
:::

::: {.column width="40%"}
-   The average year of the bridges built in Omaha, NE was `r signif(desc_brid_cond$mean[1],4)`
-   Structural condition ratings are low, as the average is shockingly below 4 (out of 10)
-   The average daily traffic (ADT_029) mean is `r format(signif(desc_brid_cond$mean[3],4),scientific = FALSE)` cars per day
:::
:::::

## Descriptive Statistics: Bridges with Poor Condition and High Traffic

::::: columns
::: {.column width="60%"}
```{r}

#| label: tbl-bridge-stats1
#| tbl-cap: "High Traffic Bridges with Poor Rating"
#| echo: false
#| message: false
#| warning: false
desc_h_t %>%
  select(mean, sd, n, median) %>%
  mutate(across(c(mean, sd, median),~ format(round(.x, 3), scientific = FALSE)))%>%
  kable() %>%
  kable_styling(font_size = 20) 

```
:::

::: {.column width="40%"}
-   The mean average daily traffic is `r format(signif(desc_h_t$mean[3],4),scientific = FALSE)`
-   The average year built is slightly newer at `r signif(desc_h_t$mean[1],4)` than bridges that are not high traffic.
-   The average number of lanes is `r signif(desc_h_t$mean[2],3)` which is much higher than the previous subset.
:::
:::::

## Descriptive Statistics: Proportion of Bridges in Poor Condition

::::: columns
::: {.column width="50%"}
```{r}
#| label: tbl-per
#| tbl-cap: "Proportion of Bridges at Risk in all of Omaha, NE"
#| echo: false
#| message: false
#| warning: false
options(digits = 4)
percents <- data.frame(
  "bridges at risk" = nrow(bridge_cond1)/nrow(fil_dat),
  "High traffic bridges at risk" = nrow(high_traffic)/nrow(fil_dat))
kable(percents, col.names = c("Bridges at Risk", "High Traffic Bridges at Risk"))
```
:::

::: {.column width="50%"}
-   The percent chance that a bridge in Omaha, NE is in poor condition is `r signif(100*percents$bridges.at.risk,3)` %.

-   The percent chance that a bridge in poor condition has high traffic is `r signif(100*percents$High.traffic.bridges.at.risk,3)`%.

-   Next the data was analyzed with a linear model
:::
:::::

## Linear Model: Collapse Risk Variable

-   A binary collapse risk variable was created
    -   It was done by creating the variable to equal 1 when the bridge condition is poor and if the structural evaluation of the bridge in less than 4, and 0 otherwise.
    -   Bridge Condition and Structural Evaluation are both summary variables

## Linear Model: Imputation Methods

-   The data set contained many "NA" values, imputation methods were used
-   Imputation methods are when you use the information that you have in the data set to estimate the missing values
-   This new data set was then used to create our linear model

## Linear Model: Model Makeup and Process

$$
\scriptsize \log\left( \frac{P(\text{collapse\_risk} = 1)}{1 - P(\text{collapse\_risk} = 1)} \right) 
= \beta_0 + \beta_1 \cdot \text{DECK\_COND\_058} \\ \scriptsize + \beta_2 \cdot \text{OPERATING\_RATING\_064} 
+ \beta_3 \cdot \text{YEAR\_BUILT\_027}
$$

-   This model is a generalized linear model
-   The response variable is non - normal (Binomial)

## Linear Model: Training and Test set

-   The data was split into a training and testing set
    -   (80% and 20% respectively)
-   This allows the model's accuracy to be tested as well as see its predictions

## Linear Model: Model Coefficients

::::: columns
::: {.column width="50%"}
```{r}
#| label: tbl-mod
#| tbl-cap: "Regression Model Summary"
#| echo: false
#| message: false
#| warning: false
options(digits = 4)
Omaha <- fil_dat %>%
  mutate(collapse_risk = ifelse(BRIDGE_CONDITION == "P" & STRUCTURAL_EVAL_067 <= 4, 1, 0)) %>%
  select(c(1,8,12,18,26,31))

library(VIM)
one <- kNN(Omaha, k = 5)


library(caret)
train_indices <- createDataPartition(one$OBJECTID, 
                                     p = 0.8, 
                                     list = FALSE)

train_data <- one[train_indices, ]

# Testing data
test_data <- one[-train_indices, ]


# Logistic regression model
model1 <- glm(collapse_risk ~as.numeric(DECK_COND_058) + OPERATING_RATING_064 + as.numeric(YEAR_BUILT_027), 
             family = binomial(link = "logit"), data = train_data)

library(gt)
library(broom)

tidy(model1) %>%
  gt()

```
:::

::: {.column width="50%"}
-   This shows that all variables are significant by having a p-value less than 0.05
-   This means that all of the coefficients will be included in the model for prediction
:::
:::::

## Linear Model: Model Accuracy

::::: columns
::: {.column width="40%"}
```{r}
#| echo: false
#| message: false
#| warning: false
options(digits = 4)
predicted_probabilities <- predict(model1, newdata = test_data, type = "response")

predicted_classes <- ifelse(predicted_probabilities > 0.5, 1, 0)

prop_pred <- mean(predicted_classes)
omaha_c <- mean(Omaha$collapse_risk)

```

```{r}
#| label: tbl-acur
#| tbl-cap: "Accuracy of Regression Model"
#| echo: false
#| message: false
#| warning: false
options(digits = 4)
# Actual values
actual_classes <- test_data$collapse_risk
 # Compare predicted vs actual values
comparison <- data.frame(Actual = actual_classes, Predicted = predicted_classes)
 
accuracy <- mean(predicted_classes == actual_classes)
kable(accuracy,col.names = c("Accuracy"))
```
:::

::: {.column width="60%"}
-   When used in prediction on the test data set, the model predicts or analyzed that there is `r signif(100*prop_pred,3)` % of bridges that are at risk of collapse
-   This is very similar to the value from the overall data which was `r signif(100*omaha_c,3)` %
-   The model has a `r signif(100*accuracy,3)`% accuracy in predicting the defined collapse risk variable
:::
:::::

# Conclusion

## Results Breakdown

-   In the analysis we have found that

    -   Bridge rating of Poor: `r signif(100*percents$bridges.at.risk,3)`%

    -   Bridge rating of poor and high traffic: `r signif(100*percents$High.traffic.bridges.at.risk,3)`%

-   Model Analysis Finding

    -   The model estimated that `r signif(100*prop_pred,3)`% of bridges in Omaha are at risk of collapse with `r signif(100*accuracy,3)`% accuracy

## Issues

-   No specifically defined criteria for if a bridge is going to collapse
    -   Lead to analyzing the descriptive statistics while also formulating a variable that categorized risk of collapse
-   Data set contained a lot of missing values
    -   The model was created using an imputed data set

## Future Implications

-   Using this model to further categorize bridges as at risk or needing attention
    -   Could allow government officials to act quicker if they know the proportion of bridges that are at risk in there district or state
    -   Lead to more economical solutions if at risk bridges can be identified then preventative measures can be taken
-   The main product of this analysis is that bridges are not in good condition and money needs to be allocated to help mend and fix bridges

## References

::: {#refs}
:::
